### 2010-03-17 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: general review

While inductive types are the alpha and omega of the simply-typed
languages, evolving in a dependent setting fosters new
opportunities. The typical example is bounded lists, also known as
vectors. A vector is a list decorated by its length. Having this
information prevents hazardous operations, such as taking the head of
an empty vector: the head function only takes vectors of length
$\NatSuc{n}$, as enforced by its type. This is made possible by the
very specificity of dependently-typed systems: terms -- the length --
are allowed to influence types -- the vector type.

@@

Before moving on indexed descriptions, we have to carry out some
maintenance work on descriptions. We presented $\Desc$ as the grammar
of inductive types. Hence, the codes closely follow this grammar. In
the following, we adopt an alternative presentation. With $\DSigma$,
we are already quoting a standard type-former. In the alternative
presentation, we go further and entirely present our codes as
quotation of standard type-formers. This presentation is shown in
Figure~\ref{fig:type-former-desc}.

@@

In the previous section, we have presented the $\Desc$ universe as the
grammar of functors in the category $\Set$. We have seen how to encode
inductive types in this setting. To encode an inductive family indexed
by $\Bhab{\V{I}}{\Set}$, we rely on functors in the category
$\Set^I$. We call these \emph{indexed functors}. When $\Desc$ was
defining a grammar a functor, we need $I \To \IDesc{I}$ to define a
grammar of indexed functors:

@@

This generalisation can be achieved by a minor evolution of $\Desc$
(Fig.~\ref{fig:factor-desc}). Indeed, the functorial nature of $\Desc$
is entirely captured by the $\DId$ code. We get an indexed functor
by turning $\DId$ into an index-dependent $\DVar$ code:

@@

The resulting code of $\IDesc$ is presented in
Figure~\ref{fig:idesc}. The reader will notice that we have also
replaced $\DUnit$ by a more general $\DConst$ code. Whereas $\DUnit$
was interpreted as the unit set, $\DConst{X}$ is interpreted as $X$,
for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
its equivalent encoding. From a definitional perspective, we are
giving more opportunities to the type-system, hence reducing the
burden on the programmer in practice.

@@

@@

In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
form for descriptions. This format follows the usual presentation of
inductive types as sums of product. With indexed description, this
definition can be generalized. Indeed, when defining an indexed
data-type, we have access to this index. Therefore, we can use this
index to influence the choice of constructors. This captures the
essence of dependent data-types: a term -- the index -- has the
ability to influence the data-type.

@@

\[\stk{
 \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{I} \spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
 \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})})  \\
 \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}}
}\]

In the case of vectors, for instance, for the index $n = \NatZero$, we
would only offer the constructor $\ListNil$. Similarly, for $n =
\NatSuc{n'}$, we only offer the constructor $\ListCons{}{}$. We will
see more instances of this pattern in the following.

@@

The function $\Val{}$, used in the definition of $\Eval{}$, simply
maps $\Epolym$ to the corresponding type in the host language. Hence,
the argument of $\Eval{}$ are ensured to be of the expected type. We
assume $\Nat$ and $\Bool$ represent natural numbers and booleans in
the host language. Further, we also assume the existence of an
addition and comparison operator in $\Nat$, respectively name
$\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$ as follow:

@@

The type of $\F{eval}$ is strikingly similar to a
catamorphism. Indeed, implementing a single step of evaluation -- the
algebra -- is sufficient, as $\F{cataI}$ gives, for free, the
full evaluator. The implementation is as follow:

@@

\note{We have not defined \F{cataI}!}

@@

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\Var{}$, it is possible to
build and manipulate \emph{open} terms, that is terms with symbolic
variables. On the model of $\Val{}$, we define $\Var{}{}$:

@@

Whereas $\Val{}$ was mapping the type to the corresponding host type,
$\Var{}{}$ maps types to a finite set. The finite set -- the context
-- contains values. A variable is therefore a $\Eval{}$ that contains
a pointer to a particular element of the finite set -- an element of
$\EnumT{dom}$. The extra argument to $\Var{}{}$ is the domain of this
context.

@@

Again, we would like to give a semantics to this extended language. We
proceed in two steps: first, replace the variables by their value in
the context; then, evaluate the resulting closed term. Thanks to
$\evalH$, we are already able to solve the second problem. Let us
focus on discharging variables from the context. Again, we can
subdivide this problem: first, we need the ability to discharge a
single variable from the context; then, we apply this $\discharge$
function on every variables in the term.

@@

Abstracting away the book-keeping introduced by the context, this
definition looks familiar. Indeed, it is extremely similar to a
monadic ``bind''. This is not surprising as we are defining a
first-order syntax with variables: our data-type enjoys more structure
than what we are currently provided. In particular, this is
reminiscent to a free monad, where $\Eval{}$ is the ``return''
introducing variables. Instead of implementing $\substH$ in this
special case, we shall implement it in a generic setting, and express
our example in that setting.

@@

However, before entering the realm of generic structures, let us
further explore the universe of indexed descriptions. In the
following, we ask $I$ to come equipped with a notion of definitional
equality. Hence, we can legibly equates elements of $I$.

@@

In our implementation of lambda terms, variables were elements of
$\Fin{n}$, with $n$ the number of binders. Let us now implement
$\Fin$. $\Fin$ is a typical example of an inductive family using
\emph{constrained constructors}. The constructors $\FinZero$ and
$\FinSuc$ are only defined for an index distinct from $\NatZero$:

@@

The $\Fin$ example is also a well-known example of Generalised
Abstract Data-Type (GADT), in Haskell. It comes as no surprise that
our universe of indexed description supports this restricted class of
dependent data-types. As we have shown with $\Fin$, constrained
constructors are straightforwardly translated into equality
constraints.

@@

Equations have simply disappeared! We should precise that forcing a
description is not guaranteed to remove all constraints. It is subject
of future work to see if constraints can be entirely eradicated, or
presented more conveniently to the user.

Finally, it is worth mentioning that the forcing optimisation is a
\emph{source-to-source} transformation of the description.

@@

In this section, we develop some more structure for our indexed
data-types. In Section~\ref{sec:desc-free-monad}, we have constructed
a free monad operation on simple descriptions. Building on this
experience, we are going to present its equivalent in the indexed
world. The processus is the same. Namely, given an indexed functor, we
derive the indexed functor coding its free monad:

@@

In Section~\ref{sec:idesc-examples}, we had the intuition that our
datatype $\HExprD$ and $\HExprVarD$ enjoys a monadic structure. We had
identified the variable substitution operation as the ``bind'' of a
free monad. The definition of $\substI$ above confirm our intuition:
using $substI$, we should easily obtain $\substH$. To do so, we first
have to massage the definition of our data-type, to exhibit its
monadic structure.

@@

The resulting data-type is called $\HExprFreeD$. By a simple unfolding
of definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$
corresponds to the closed expression, $\HExprD$. Similarly,
$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{dom}{ty})}$$
corresponds to expressions with variables, $\HExprVarD$.

@@

The evaluator for closed term we implemented in
Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$ to values in
$\Val{ty}$. We are therefore left with implementing $\substH$. We
simply have to fill in the right arguments to $\substI$, the type
guiding us:

@@

Another instance of free monad is $\IDesc$ itself. Indeed, $\DVar$ is
nothing but the ``return''. The remaining constructors are simply the
freed functor, trivially indexed by $\Unit$. Hence, amusingly, we can
define the $\IDesc$ freed functor by the following code:

>>>>

While inductive types are the alpha and omega of simply-typed
languages, evolving in a dependent setting fosters new
opportunities. The typical example is bounded lists, also known as
vectors. A vector is a list decorated by its length. Having this
information prevents hazardous operations, such as taking the head of
an empty vector: the head function only takes vectors of length
$\NatSuc{n}$, as enforced by its type. This is made possible by the
very specificity of dependently-typed systems: terms -- the length --
are allowed to influence types -- the vector type.

@@

Before moving on indexed descriptions, we have to carry out some
maintenance work on descriptions. We presented $\Desc$ as the grammar
of inductive types. Hence, the codes closely follow this grammar. In
the following, we adopt an alternative presentation. With $\DSigma$,
we are actually \emph{quoting} a standard type-former, namely

$$\Bhab{\Sigma}{\PI{\V{S}}{\Set} (S \To \Set) \To \Set}$$

In the alternative presentation, we go further and present all our
codes as quotations of standard type-formers. This presentation is
shown in Figure~\ref{fig:type-former-desc}.

@@

In the previous section, we have presented the $\Desc$ universe as the
grammar of functors in the category $\Set$. We have seen how to encode
inductive types in this setting. To encode an inductive family indexed
by $\Bhab{\V{I}}{\Set}$, we rely on functors in the category
$\Set^I$. We call these \emph{indexed functors}. In the following, we
implement $I \To \IDesc{I}$ as a grammar for indexed functors:

@@

This generalisation can be achieved by a minor evolution of $\Desc$
(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of $\Desc$
is entirely captured by the $\DId$ code. We get an indexed functor
by turning $\DId$ into an index-dependent $\DVar$ code:

@@

The resulting code of $\IDesc$ is presented in
Figure~\ref{fig:idesc}. The reader will notice that we have also
replaced $\DUnit$ by a more general $\DConst$ code. Whereas $\DUnit$
was interpreted as the unit set, $\DConst{X}$ is interpreted as $X$,
for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
its equivalent encoding. From a definitional perspective, we are
giving more opportunities to the type-system, hence reducing the
burden on the programmer.

@@

\subsection{Tagged indexed descriptions}

@@

In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
form for descriptions. This format follows the usual presentation of
inductive types as sum-of-product. With indexed description, this
definition can be generalized. Indeed, when defining an indexed
data-type, we have access to this index. Therefore, we can use this
index to influence the choice of constructors. This captures the
essence of dependent data-types: a term -- the index -- has the
ability to influence the data-type.

@@

\[\stk{
 \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}}
 \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{I} \spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
 \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})})  \\
}\]

In the case of vectors, for instance, for the index $\NatZero$, we
would only propose the constructor $\ListNil$. Similarly, for
$\NatSuc{n}$, we only propose the constructor $\ListCons{}{}$. We will
see more instances of this pattern in the following.

@@

The function $\Val{}$, used in the definition of $\Eval{}$, simply
maps $\Epolym$ to the corresponding type in the host language. Hence,
the argument of $\Eval{}$ are ensured to be of the expected type. We
assume $\Nat$ and $\Bool$ represent natural numbers and booleans in
the host language. Further, we also assume the existence of an
addition and comparison operator in $\Nat$, respectively named
$\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$ as follow:

@@

The type of $\F{eval}$ is strikingly similar to a
catamorphism. Indeed, implementing a single step of evaluation -- the
algebra -- is sufficient, as $\F{cataI}$ \note{We have not defined
  \F{cataI}!} gives, for free, the full evaluator. The implementation
is as follow:

@@

@@

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\Val{}$, it is possible to
build and manipulate \emph{open} terms, that is terms with symbolic
variables. On the model of $\Val{}$, we define $\Var{}{}$:

@@

Whereas $\Val{}$ was mapping the type to the corresponding host type,
$\Var{}{}$ maps types to a finite set. The finite set -- the context
-- contains closed terms. A variable is therefore a $\Eval{}$ that
contains a pointer to a particular element of the finite set -- an
element of $\EnumT{dom}$. The extra argument to $\Var{}{}$ is the
domain of this context.

@@

Again, we would like to give a semantics to this extended language. We
proceed in two steps: first, we replace the variables by their value
in the context; then, we evaluate the resulting closed term. Thanks to
$\evalH$, we are already able to solve the second problem. Let us
focus on discharging variables from the context. Again, we can
subdivide this problem: first, we need the ability to discharge a
single variable from the context; then, we apply this $\discharge$
function on every variables in the term.

@@

Abstracting away the book-keeping introduced by the context, this
definition looks familiar. Indeed, it is extremely similar to a
monadic ``bind''. This is not surprising as we are defining a
first-order syntax with variables: our data-type enjoys more structure
than what we are given. In particular, this is reminiscent to a free
monad, where $\Eval{}$ is the ``return'' introducing
variables. Instead of implementing $\substH$ in this special case, we
shall implement it in a generic setting, and express our example in
that setting. However, before entering the realm of generic
structures, let us further explore the universe of indexed
descriptions.

@@

In the following, we ask $I$ to come equipped with a notion of
definitional equality. Hence, we can legibly equates elements of $I$.

@@

Another common dependent data-type is $\Fin$, the finite sets of
numbers. $\Fin$ is a typical example of an inductive family using
\emph{constrained constructors}. The constructors $\FinZero$ and
$\FinSuc$ are only defined for an index distinct from $\NatZero$:

@@

The $\Fin$ example is also a well-known example of Generalised
Abstract Data-Type (GADT) in Haskell. It comes as no surprise that our
universe of indexed description supports this restricted class of
dependent data-types. As we have shown with $\Fin$, constrained
constructors are straightforwardly translated into equality
constraints.

@@

Equations have simply disappeared! We should precise that forcing a
description is not guaranteed to remove all constraints. It is subject
of future work to see if constraints can be entirely eradicated, or
presented more conveniently to the user. Finally, it is worth
mentioning that the forcing optimisation is a \emph{source-to-source}
transformation of the description.

@@

In this section, we develop more structure for our indexed
data-types. In Section~\ref{sec:desc-free-monad}, we have constructed
a free monad operation on simple descriptions. Building on this
experience, we are going to present its equivalent in the indexed
world. The processus is the same. Namely, given an indexed functor, we
derive the indexed functor coding its free monad:

@@

In Section~\ref{sec:idesc-examples}, we had the intuition that our
datatypes $\HExprD$ and $\HExprVarD$ enjoy a monadic structure. We had
identified the variable substitution operation as the ``bind'' of a
free monad. The definition of $\substI$ above confirm our intuition:
using $\substI$, we should easily obtain $\substH$. To do so, we first
have to massage the definition of our data-type, to exhibit its
monadic structure.

@@

The resulting data-type is called $\HExprFreeD$. By a simple unfolding
of definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$
corresponds to the syntax of closed terms, $\HExprD$. Similarly,
$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{dom}{ty})}$$
corresponds to expressions with variables, $\HExprVarD$.

@@

The evaluator for closed term we implemented in
Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}\: ty$ to values in
$\Val{ty}$. We are therefore left with implementing $\substH$. We
simply have to fill in the right arguments to $\substI$, the type
guiding us:

@@

Another instance of free monad is $\IDesc$ itself. Indeed, $\DVar$ is
nothing but the ``return''. The remaining constructors are simply the
signature functor, trivially indexed by $\Unit$. Hence, we describe
this signature functor by the following code:

### 2010-03-17 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: general review

Noticeably, we have adopted a judgemental presentation of
equality. Therefore, we are not tied to a particular implementation
choice. In particular, our system has been modelled in Agda, which
features an in-tensional equality. On the other hand, it is used in
Epigram, which equality has a slightly extensional
flavor~\cite{altenkirch:ott}. We expect users of fully extensional
systems to also find their way through this presentation.

@@

Finally, we define the rules governing judgemental equality in
Figure~\ref{fig:judgemental-equality}. As one would expect,
judgemental equality respects symmetry, reflexivity, and
transitivity. Further, it encompasses the $\beta$-rules for function
application and pairs. Finally, it respects structural equality,
including equality under lambda ($\xi$-rule).

Crucially, being judgemental, this presentation is agnostic in the
notion of equality actually implemented. Indeed, our typing and
equality judgements do not rely on a ``propositional'' equality. This
freedom is a key point in favour of the wide applicability of our
proposal. Obviously, when implementing our technology in an existing
type-theory, some opportunities arise. We will present some of them
along the course of the paper. \note{It is just a specification. We
  expect the propositional theory to provide a decidable machinery to
  implement it.}

\note{plus structural rules, including lambda ($\xi$-rule)}

@@

The specification of finite sets is presented in
Figure~\ref{fig:typing-finite-set}. It is composed by three
components. First, we define tags as inhabitants of the $\UId$ type. A
tag is solely an informative token, used for display purposes. Finite
sets inhabits the $\EnumU$ type. Unfolding the definition, we get that
a finite set is a list of tags. Finally, elements of a finite set $u$
belong to the corresponding $\EnumT{u}$ type. Intuitively, it
corresponds to an index, a number, pointing to an element of $u$.

@@

\note{Needs to be defined.}

@@

Our proposal to tame the problem of codes is deeply rooted in the
bidirectional presentation of type-checking of Pierce and
Turner~\cite{pierce:bidirectional-tc}. In this formalism,
type-checking is divided into two complementary phases. In the
\emph{type inference} phase, types are \emph{pulled} out of terms. A
typical example is a variable in the context:

@@

We adapt and formalise this \emph{type propagation} mechanism to our
richer type theory. First of all, we define a high-level syntax,
composed by \emph{expressions}. Expressions divided into two syntactic
categories: $\exprIn$ into which types are pushed in, and $\exprEx$
from which types are pulled out. Following the bidirectional
presentation, $\exprIn$ are subject to type \emph{checking}, while
$\exprEx$ are subject to type \emph{synthesis}.

@@

The purpose of type propagation is to \emph{elaborate} an high-level
expression into a term of the type theory, together with its
type. This presentation builds on the experience gained in
implementing Epigram 1~\cite{mcbride:view-from-the-left}. Being
entirely syntax driven, we present our formalism as a system of
syntax-directed inference rules. As the presentation largely mirrors
the inference rules of the type theory, we will ignore the judgements
which are identical. We refer our reader to the associated technical
report~\cite{chapman:desc-tech-report} for the complete system of
rules.

@@

The type synthesis rules are presented
Figure~\ref{fig:type-synthesis}. As illustrated by the judgement form,
a type synthesis problem is provided an expression $\exprEx$ in a
context, which is elaborated into a low-level term together with its
type. These terms act as \emph{sources} for the type propagation
machinery: they make types out of expressions. We have already studied
the case of a variable in context. This also include function
application as well as projections. Further down the line, these types
might propagate to other terms, thanks to type checking.

@@

A second consequence concerns the definition of a finite set. We can
abstract away the details of code and use a familiar syntax
instead. Hence, we use the syntax $\Enum\: [t_1 \ldots t_n]$ to denote
the finite sets composed by the named elements $t_1$, \ldots,
$t_n$. The translation to elements of $\EnumU$ is straightforward.

@@

The abstraction we have developed in this section are entirely
transparent to the user. In the case of finite function space, the
user has never to explicitly \emph{call} the $\F{switch}$
eliminator. For diagnosis purposes, a similar, reverse mechanism turns
low-level terms into high-level expressions. Hence, this system, which
presentation is out of the scope of this paper, seals the low-level
type theory behind the high-level expressions.

@@

In this section, we have developed a type propagation system, based
on bidirectional type-checking. Using bidirectional type-checking as a
boilerplate scrapper is a well-known technique~\cite{who?} \note{I
  would need some citation of people using bidirectional presentation
  of their type theory}. In our case, we have shown how to instrument
bi-directionality to rationalise the expressivity of dependent
types. We have illustrated our approach with finite sets. We have
successfully abstracted away the low-level presentation of finite
sets, offering a convenient syntax instead.

>>>>

Noticeably, we have adopted a judgemental presentation of
equality. Therefore, we are not tied to a particular implementation
choice. In particular, our system has been modelled in Agda, which
features an intentional equality. On the other hand, it is used in
Epigram, which equality has a slightly extensional
flavor~\cite{altenkirch:ott}. We expect users of fully extensional
systems to also find their way through this presentation.

@@

Finally, we define the rules governing judgemental equality in
Figure~\ref{fig:judgemental-equality}. We implicitly assume that
judgemental equality respects symmetry, reflexivity, and
transitivity. We capture the computational behavior of the language
through the $\beta$-rules for function application and pairs. Finally,
we implicitly assume that it respects purely syntactic and structural
equality. This includes equality under lambda ($\xi$-rule).

Crucially, being judgemental, this presentation is agnostic in the
notion of equality actually implemented. Indeed, our typing and
equality judgements do not rely on a ``propositional'' equality. This
freedom is a key point in favour of the wide applicability of our
proposal. This judgemental presentation must be read as a
\emph{specification}: our proposal works with any propositional
equality satisfying this specification. Moreover, our lightweight
requirements do not endanger decidability of equality-checking.
Obviously, when implementing our technology in an existing
type-theory, some opportunities arise. We will present some of them
along the course of the paper.

@@

The specification of finite sets is presented in
Figure~\ref{fig:typing-finite-set}. It is composed of three
components. First, we define tags as inhabitants of the $\UId$ type. A
tag is solely an informative token, used for diagnostic
purposes. Finite sets inhabits the $\EnumU$ type. Unfolding the
definition, we get that a finite set is a list of tags. Finally,
elements of a finite set $u$ belong to the corresponding $\EnumT{u}$
type. Intuitively, it corresponds to an index -- a number -- pointing
to an element of $u$.

@@

@@

Our proposal to tame the problem of codes is deeply rooted in the
bidirectional presentation of type-checking from Pierce and
Turner~\cite{pierce:bidirectional-tc}. In this formalism,
type-checking is divided into two complementary phases. In the
\emph{type inference} phase, types are \emph{pulled} out of terms. A
typical example is a variable in the context:

@@

We adapt and formalise this \emph{type propagation} mechanism to our
richer type theory. First of all, we define a high-level syntax,
composed by \emph{expressions}. Expressions are divided into two
syntactic categories: $\exprIn$ into which types are pushed in, and
$\exprEx$ from which types are pulled out. Following the bidirectional
presentation, $\exprIn$ are subject to type \emph{checking}, while
$\exprEx$ are subject to type \emph{synthesis}.

@@

The purpose of type propagation is to \emph{elaborate} an high-level
expression into a term of the type theory, together with its
type. This presentation builds on the experience gained in
implementing Epigram 1~\cite{mcbride:view-from-the-left}. Being
entirely syntax driven, we present our formalism as a system of
syntax-directed inference rules. As the presentation largely mirrors
the inference rules of the type theory, we will ignore the judgements
that are identical. We refer our reader to the associated technical
report~\cite{chapman:desc-tech-report} for the complete system of
rules.

@@

The type synthesis rules are presented
Figure~\ref{fig:type-synthesis}. As illustrated by the judgement form,
a type synthesis problem is provided an expression $\exprEx$ in a
context, which is elaborated into a low-level term together with its
type. These terms act as \emph{sources} for the type propagation
machinery: they make types out of expressions. We have already studied
the case of a variable in context. This also include function
application as well as projections. Further down the line, these
inferred types might propagate to other terms, thanks to type
checking.

@@

A second consequence concerns the definition of a finite set. We can
abstract away the details of code and use a familiar syntax
instead. Hence, we use the syntax $[t_1 \ldots t_n]$ to denote the
finite sets composed by the named elements $t_1$, \ldots, $t_n$. The
translation to elements of $\EnumU$ is straightforward.

@@

The abstraction we have developed in this section are entirely
transparent to the user. In the case of finite function space, the
user has never to explicitly \emph{call} the $\F{switch}$
eliminator. For diagnosis purposes, a similar, reverse mechanism turns
low-level terms into high-level expressions. Hence, this system, which
complete presentation is out of the scope of this paper, seals the
low-level type theory behind the high-level expressions.

@@

In this section, we have developed a type propagation system based on
bidirectional type-checking. Using bidirectional type-checking as a
boilerplate scrapper is a well-known technique~\cite{who?} \note{I
  would need some citation of people using bidirectional presentation
  of their type theory}. In our case, we have shown how to instrument
bi-directionality to rationalise the expressivity of dependent
types. We have illustrated our approach with finite sets. We have
abstracted away the low-level presentation of finite sets, offering a
convenient syntax instead.
